On investigating heterogeneity in psychological phenomena

### 2023-10-30 ms.Rmd
WIP
First draft of complete analyses.

---
title: 'Bayesian assessment of heterogeneity in psychological phenomena'
shorttitle: Heterogeneity in psychology
leftheader: Heterogeneity in psychology
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Tilburg School of Social and Behavioral Sciences, Tilburg University
    email: m.j.vuorre@tilburguniversity.edu
  - name: Matthew Kay
    affiliation: 2
  - name: Niall Bolger
    affiliation: 3
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg University
  - id: 2
    institution: Computer Science and Communication Studies, Northwestern University
  - id: 3
    institution: Department of Psychology, Columbia University
abstract: |
  Advances in statistics and data collection methods have brought variability in psychological phenomena to the forefront of theoretical development. Yet, established methods and practices for describing heterogeneity underestimate the true uncertainty in those estimates. In the current work, we aim to deliver methods and guidelines for more rigorous treatments of heterogeneity in psychological phenomena.
  We provide a walkthrough of methods for i. estimating, ii. summarising, and iii. describing heterogeneity. Although other methods for incorporating estimation uncertainty are available, the methods we propose---involving bayesian hierarchical models---are especially suitable for investigating heterogeneity because they naturally describe uncertainty in all model parameters.
  We hope that the proposed methods help researchers to widen their inferential focus from statistical averages to include statistical heterogeneity in psychological phenomena.
keywords: variation, heterogeneity, hierarchical models, mixed models, methodology
authornote: |
  \noindent \textbf{This working paper is not yet peer reviewed.}
wordcount: "`r wordcountaddin:::word_count(filename = 'ms.Rmd')`"
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{4pt}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

```{r setup, include = FALSE}
# Packages
library(papaja)
library(janitor)
library(scales)
library(ellipse)
library(knitr)
library(latex2exp)
library(lme4)
library(ggpp)
library(patchwork)
library(latex2exp)
library(ggdist)
library(ggstance)
library(tidybayes)
library(distributional)
library(parameters)
library(brms)
library(tidyverse)

@@

# Plotting options
theme_set(
  theme_classic(base_size = 9) +
    theme(
      strip.text = element_text(color = "black", hjust = 0),
      panel.grid = element_blank(),
      line = element_line(linewidth = .25)
    )
)

# Model estimation options
dir.create("models", FALSE)
options(
  mc.cores = 4,
  brms.backend = ifelse(require(cmdstanr), "cmdstanr", "rstan")
)

dir.create("data", FALSE)
```

@@

One reason for the scarcity of reporting and interpreting heterogeneity is that psychologists still tend to analyze data with models that obscure the assessment of heterogeneity, such as ANOVA [@bolgerCausalProcessesPsychology2019]. Although more appropriate hierarchical (or multilevel, mixed-effects) models are rapidly becoming more widespread, we believe a second important reason is that researchers do not yet have the conceptual and practical tools required to estimate and describe heterogeneity. In the current work, we propose several practical methods, descriptions, and visualizations that allow researchers to better estimate, interpret, and report heterogeneity in a manner that maximizes empirical results' impact on theory building.

@@

# Save variables as numerical & factor (might need for more complex contrast/eff/etc codes)
dat <- read_csv("data/bolger-etal/heterogeneity_dataset1_traitvalence.csv") |> 
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) |> 
  # Some more cleaning
  mutate(
    person = fct_anon(factor(id)),
    trial,
    valence = factor(valenceE, levels = c(-0.5, 0.5), labels = c("Negative", "Positive")),
    rt = logrt
  ) |> 
  select(person, trial, valence, rt) |> 
  arrange(person, trial)
```

```{r contrast-coding, include = FALSE}
contrasts(dat$valence) <- c(-0.5, 0.5)
contrasts(dat$valence)
```

@@

```{r}
#| label: tab-dat1
dat |> 
  head() |> 
  apa_table(
    span_text_columns = FALSE, 
    digits = c(0,0,0,2),
    caption = "First six rows of example dataset 1 (Bolger et al., 2019; Exp 1).", 
    note = "rt is log-transformed."
  )
```

@@

The lack of subscripts on $\sigma$ indicates that we assume it to be invariant across trials and subjects. Then, we specify a model of the mean $\eta_{ij}$ such that the regression coefficients capture our substantive questions:

@@

We contrast coded valence such that negative words are coded as -0.5, and positive words as 0.5. This coding results in an intercept that corresponds to the average reaction time across negative and positive words, and a slope coefficient that reflects the difference in log(rt) between negative and positive words. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$ are subject-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the average reaction time for subject $j=1$. The $\gamma$ coefficients have subscripts, which in this context indicates that they are further modelled on covariates---here, index variables for "person" in the data. We refer to these as "group level" parameters, in line with conventions. In the frequentist discourse, these parameters are called "random" parameters [@gelmanDataAnalysisUsing2007]. Again, readers can equate the term "group level" parameters with "random" parameters in the ongoing discussion.

@@

```{r fit1-lmer}
fit1_lmer <- lmer(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat
)
```

```{r}
# Create a table of coefficients
tab1_lmer <- model_parameters(fit1_lmer, effects = "all") |> 
  tibble() |> 
  select(Parameter, Coefficient, starts_with("CI_"))

# Bootstrap CIs for variance parameters
tab1_lmer[3:6,3:4] <- confint(fit1_lmer, method = "boot", nsim = 100)[c(1,3,2,4),1:2]
```

```{r fit1-brms}
fit1 <- brm(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat,
  file = "models/brm-fit-1",
  control = list(adapt_delta = .95)
)
```

tab1_brms <- parameters(fit1, centrality = "mean", effects = "all") |> 
  tibble() |> 
  select(Parameter, Mean, starts_with("CI_"))
```

@@

tab1_lmer |> 
  mutate(
    across(where(is.numeric), ~number(., .01)),
    MLE = str_glue("{Coefficient} [{CI_low}, {CI_high}]"),
  ) |> 
  select(Parameter, `Estimate [95\\%CI]` = MLE) |> 
  mutate(
    Parameter = m1_parnames
  ) |> 
  apa_table(
    span_text_columns = FALSE,
    caption = "Parameter estimates from Model 1.",
    escape = FALSE
  )
mu1 <- number(as.numeric(tab1_lmer[2,2:4,T]))
sd1 <- number(as.numeric(tab1_lmer[4,2:4,T]))
mu1n <- tab1_lmer[2,2,TRUE]
sd1n <- tab1_lmer[4,2,TRUE]
```

@@

```{r descriptors, include = FALSE}
# Heterogeneity interval
hi90_num <- mu1n + qnorm(c(0.05, .95)) * sd1n
hi90 <- paste(
  number(hi90_num, .01), 
  collapse = ", "
)
hi90_low <- paste(
  number(tab1_lmer[2,3,T] + qnorm(c(0.05, .95))*tab1_lmer[4,3,T], .01), 
  collapse = ", "
)

@@

```{r}
#| label: hdist-1
#| fig.height: 2.2
#| fig.width: 8
#| fig.env: "figure*"
#| fig.cap: The heterogeneity distribution and various descriptions of the expected heterogeneity in valence effects from Model 1. **A**. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. **B**. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. **C**. Proportion of negative valence effects in the population (dark). **D**. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

p0 <- ggplot() +
  aes(xdist = dist_normal(mu1n, sd1n)) +
  scale_x_continuous(breaks = extended_breaks()) +
  coord_cartesian(xlim = c(-0.55, 0.25)) +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(lineheight = rel(0.5))
    ) +
  scale_fill_manual(values = c("grey90", "grey80"), na.value = "white")
d1 <- dist_normal(mu1n, sd1n) |> 
  mean_qi(.width = c(.66, .9))

p1 <- p0 +
  stat_slab(
    aes(fill = after_stat(level)), 
    .width = c(.68, .95), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$\\tau_1$"),
    x = mu1n + .05, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n + sd1n, y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  geom_text(
    label = TeX("$\\beta_1$"),
    x = mu1n-.05, y = .075,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n, y = .1, yend = 0,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  theme(axis.title.y = element_text(angle = 90))

p2 <- p0 + 
  stat_slab(
    aes(fill = after_stat(level)), 
    .width = c(.90, 1), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$HI_{90}$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = d1$.lower[2], xend = d1$.upper[2], y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed", ends = "both")
  )

p3 <- p0 + 
  stat_slab(
    aes(fill = after_stat(x < 0)), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX('$p^d$'),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  )

p4 <- p0 + 
  stat_slab(
    aes(fill = after_stat(between(x, -.1, .1))), 
    p_limits = c(.000001, .999999), 
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX('$p^{ROPE}$'),
    x = 0, y = .07,
    size = 3,
    parse = TRUE
  )

(p1 | p2 | p3 | p4) &
  scale_y_continuous(
    expand = expansion(c(0.0, 0.05))
  ) &
  labs(
    y = "Density",
    x = "Valence effect"
  ) &
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  ) &
  plot_annotation(tag_levels = "A")
```

The point estimates in Table \@ref(tab:tbl-fit-1) define the expected normal distribution of valence effects in the population, visualized in Figure \@ref(fig:hdist-1)A. As such, Normal(`r mu1`, `r sd1`^2^) is a valid description of heterogeneity in valence effects. However, it is an incomplete description of heterogeneity for two reasons. First, it does not necessarily communicate the degree of heterogeneity in clear and actionable terms. Second, it ignores the uncertainty with which the heterogeneity parameters are estimated.

@@

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use the .05 and .95 percentiles (which define the central 90% of the distribution). Then, we pass those percentiles and the estimated mean and standard deviation (see Table \@ref(tab:tbl-fit-1)) to the normal quantile function ($q$; `qnorm()` in R) to obtain the interval: $HI_{90} = q([.05, .95], \beta_1, \tau_1) = q([.05, .95],$ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this equation calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall within this interval. We illustrate this interval in Figure \@ref(fig:hdist-1)B.

@@

For many applications, quantities such as proportions of slopes above or below some critical value, or within some critical range, might serve as more informative summaries. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" Thus, we first define a critical value, here zero. We then pass that value and the valence effect's estimated mean and standard deviation to the cumulative normal distribution function ($p$; `pnorm()` in R) to obtain the probability density in this distribution that falls below zero: $p^d = p(0, \beta_1, \tau_1) = p(0,$ `r mu1`, `r sd1`) = `r p_less_zero`. Specifically, this metric indicates the proportion of individuals in the population with negative effects. In words, we have estimated that `r p_less_zero` of the population respond faster to positively valenced words. We show this area under the curve in Figure \@ref(fig:hdist-1)C.

However, using zero as a critical value might not be sufficiently informative especially when theory allows defining a smallest effect sizes of interest, or what is known as a region of practical equivalence (to zero) [@anvariUsingAnchorbasedMethods2021; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014]. For example, we might know from theory that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the cumulative normal distribution function ($p$) to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = p(0.1, \beta_1, \tau_1) - p(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero (within the [-0.1, 0.1] interval). We visualize this area under the curve in Figure \@ref(fig:hdist-1)D.

@@

Although informative, the expected normal distribution of valence effects, or any transformation of those point estimates such as the ones illustrated in Figure \@ref(fig:hdist-1), ignores the uncertainty inherent in the estimated parameters, and therefore cannot communicate the analyst's uncertainty about the degree of heterogeneity. That is, the uncertainty with which we have estimated the population parameters, indicated by e.g. the confidence intervals in Table \@ref(tab:tbl-fit-1), is not carried forward to the calculations described above. For example, we only presented the expected heterogeneity interval, but not the range of credible values that it might take. For instance, if we re-calculate $HI_{90}$ based on the lowest 95%CI bounds in Table \@ref(tab:tbl-fit-1), the result is a heterogeneity interval that indicates 90% of the population's slopes falling in the [`r hi90_low`] interval. In contrast to the first interval, this interval implies much less heterogeneity and that 90% of the population's slopes are expected to fall below zero.

@@

We think that probabilistic (i.e. "bayesian") methods are uniquely applicable to estimating and describing heterogeneity in psychological phenomena in ways that naturally incorporate information about uncertainty because they deliver distributions of plausible values that could underlie hypothetical data generating processes [@gelmanBayesianDataAnalysis2013; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014; @mcelreathStatisticalRethinkingBayesian2020]. Moreover, these methods are widely available in modern software packages; here we used the R package brms, which uses familiar R formulas for specifying bayesian models [@burknerAdvancedBayesianMultilevel2018; @burknerBrmsPackageBayesian2017; @rcoreteamLanguageEnvironmentStatistical2023].

@@

post1 <- as.data.frame(fit1, variable = c("b_", "sd_", "cor_", "sigma"), regex = TRUE) |> 
  tibble() |> 
  rownames_to_column("Sample")
post1_sum <- fit1 |> 
  parameters(effects = "all", centrality = "mean", dispersion = TRUE) |> 
  tibble() |> 
  mutate(
    Mean = number(Mean, .01), 
    SD = number(SD, .001)
  ) |> 
  select(Parameter, Mean, SD) |> 
  pivot_longer(c(Mean, SD), names_to = "Sample") |>
  pivot_wider(names_from = Parameter, values_from = value)

set.seed(191)
post1 |> 
  slice_sample(n = 6) |> 
  arrange(as.integer(Sample)) |> 
  mutate(
    across(-Sample, ~number(., .01))
  ) |>
  add_row(post1_sum) |> 
  setNames(c("Sample", m1_parnames)) |> 
  apa_table(
    span_text_columns = FALSE,
    font_size = "small",
    midrules = 6,
    caption = "Six random samples from Model 1's parameters' posterior distributions, and their summaries.", 
    escape = FALSE,
    placement = "h"
  )
```

The goal of bayesian inference is the multivariate posterior probability distribution of the model's parameters. However, because closed-form solutions are not available for posterior distributions of many important types of statistical models, in practice bayesian methods rely on drawing many random samples from the posterior distribution [@gelmanBayesianDataAnalysis2013; @ravenzwaaijSimpleIntroductionMarkov2016]. Therefore, "estimate" means a large number of random samples from the posterior distribution. These samples can then be used to calculate and summarize the posterior distribution of any desired quantity.

>>>>

---
title: 'On investigating heterogeneity in psychological phenomena'
shorttitle: Heterogeneity in psychology
leftheader: Heterogeneity in psychology
author: 
  - name: Matti Vuorre
    affiliation: 1
    corresponding: yes
    address: Tilburg School of Social and Behavioral Sciences, Tilburg University
    email: mjvuorre@uvt.nl
  - name: Matthew Kay
    affiliation: 2
  - name: Niall Bolger
    affiliation: 3
affiliation:
  - id: 1
    institution: Department of Social Psychology, Tilburg School of Social and Behavioral Sciences, Tilburg University
  - id: 2
    institution: Computer Science and Communication Studies, Northwestern University
  - id: 3
    institution: Department of Psychology, Columbia University
abstract: |
  Advances in statistics and data collection methods have brought variability in psychological phenomena to the forefront of theoretical development. Yet, established methods and practices for studying heterogeneity underestimate the true uncertainty in those estimates. In the current work, we aim to deliver methods and guidelines for more rigorous treatments of heterogeneity in psychological phenomena. We provide a walkthrough of methods for i. estimating, ii. summarising, and iii. describing heterogeneity. Although other methods for incorporating estimation uncertainty are available, the methods we propose---involving bayesian hierarchical models---are especially suitable for investigating heterogeneity because they naturally describe uncertainty in all model parameters. We hope that the methods we discuss would help researchers to widen their inferential focus from statistical averages to include statistical heterogeneity in psychological phenomena.
keywords: variation, heterogeneity, hierarchical models, mixed models, methodology
authornote: |
  \noindent \textbf{This working paper is not yet peer reviewed.}
wordcount: 5196
bibliography: references.bib
floatsintext: yes
numbersections: false
linenumbers: no
draft: no
mask: no
figurelist: no
tablelist: no
footnotelist: no
documentclass: apa7
classoption: jou
header-includes:
  - |
    \usepackage{float}
    \setlength{\parskip}{4pt}
output: 
  papaja::apa6_pdf:
    number_sections: true
    keep_tex: false
    highlight: kate
  papaja::apa6_docx: 
    number_sections: true
    keep_tex: false
---

```{r setup, include = FALSE}
# Packages
library(papaja)
library(janitor)
library(scales)
library(ellipse)
library(knitr)
library(readxl)
library(latex2exp)
library(lme4)
library(ggpp)
library(patchwork)
library(latex2exp)
library(ggdist)
library(ggstance)
library(tidybayes)
library(distributional)
library(posterior)
library(parameters)
library(brms)
library(tidyverse)

@@

# Plotting options
theme_set(
  theme_classic(base_size = 9) +
    theme(
      strip.text = element_text(color = "black", hjust = 0),
      strip.background = element_rect(color = NA, fill = NA),
      strip.text.x = element_text(size = rel(0.9)),
      line = element_line(linewidth = .25),
      plot.tag = element_text(size = rel(1))
    )
)

# Model estimation options
dir.create("models", FALSE)
options(
  mc.cores = 4
)
if (require(cmdstanr)) {
  options(
    brms.backend = "cmdstanr",
    brms.threads = 2
  )
}  
dir.create("data", FALSE)
```

@@

One reason for the scarcity of reporting and interpreting heterogeneity is that psychologists still tend to analyze data with models that obscure the assessment of heterogeneity, such as ANOVA [@bolgerCausalProcessesPsychology2019]. Although more appropriate hierarchical (or multilevel, mixed-effects) models are rapidly becoming more widespread, we believe a second important reason is that researchers do not yet have the conceptual and practical tools required to sufficiently estimate and describe heterogeneity. In the current work, we highlight several practical methods, descriptions, and visualizations that allow researchers to better estimate, interpret, and report heterogeneity in a manner that maximizes empirical results' impact on theory building.

@@

# Save variables as numerical & factor
dat <- read_csv("data/bolger-etal/heterogeneity_dataset1_traitvalence.csv") |>
  # Cleaning as in Bolger et al 2019
  filter(
    response.keys == "up",
    scale(rt) < 3,
    !(id %in% c(250, 257, 272))
  ) |>
  # Some more cleaning
  mutate(
    person = factor(id),
    trial,
    valence = factor(
      valenceE,
      levels = c(-0.5, 0.5),
      labels = c("Negative", "Positive")
    ),
    rt = logrt
  ) |>
  select(person, trial, valence, rt) |>
  arrange(person, trial)

contrasts(dat$valence) <- c(-0.5, 0.5)
contrasts(dat$valence)
```

@@

```{r}
#| label: tab-dat1
dat |>
  head() |>
  apa_table(
    span_text_columns = FALSE,
    font_size = "small",
    digits = c(0, 0, 0, 2),
    caption = "First six rows of example dataset 1 (Bolger et al., 2019; Exp 1).",
    note = "rt is log-transformed."
  )
```

@@

The lack of subscripts on $\sigma$ indicates that we assume it to be invariant across trials and subjects. (Note that there are better alternatives for modelling reaction times (RT) than modelling the log-transformed RTs as normal, but those are outside the scope of this manuscript.) Then, we specify a model of the mean $\eta_{ij}$ such that the regression coefficients capture our substantive questions:

@@

We contrast coded valence such that negative words are coded as -0.5, and positive words as 0.5. This coding results in an intercept that corresponds to the average reaction time across negative and positive words, and a slope coefficient that reflects the difference in log(rt) between negative and positive words. The second set of parameters, $\gamma_{0j}$ (*gamma*) and $\gamma_{1j}$ are person-specific deviations from the average intercept and slope, respectively. That is, $\beta_0 + \gamma_{01}$ is the average reaction time for person $j=1$. The $\gamma$ coefficients have subscripts, which in this context indicates that they are further modelled on covariates---here, index variables for "person" in the data. We refer to these as "group level" parameters, in line with conventions. In the frequentist discourse, these parameters are called "random" parameters [@gelmanDataAnalysisUsing2007]. Again, readers can equate the term "group level" parameters with "random" parameters in the ongoing discussion.

@@

```{r fit1-lmer}
fit_lmer <- lmer(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat
)
```

```{r}
# Create a table of coefficients
tab_1 <- model_parameters(fit_lmer, effects = "all") |>
  tibble() |>
  select(Parameter, Coefficient, starts_with("CI_"))

# Bootstrap CIs for variance parameters
tab_1[3:6, 3:4] <- confint(
  fit_lmer,
  method = "boot",
  nsim = 100,
  oldNames = FALSE
)[c(1, 3, 2, 4), 1:2]

# Get approx SEs
tab_1 <- tab_1 |> 
  mutate(SE = (CI_high - Coefficient) / 1.96) |> 
  select(1, 2, 5, 3:4)
```

@@

tab_1 |>
  mutate(
    mutate(across(where(is.numeric), ~number(.x, .01))),
    CI = str_glue("[{CI_low}, {CI_high}]"),
    Coefficient = ifelse(
      str_starts(Coefficient, "-"), 
      Coefficient, 
      paste0("\\phantom{-}", Coefficient)
    ),
    Parameter = m1_parnames
  ) |> 
  select(
    Parameter, Coefficient, SE, CI
  ) |>
  apa_table(
    span_text_columns = FALSE,
    font_size = "small",
    caption = "Parameter estimates from Model 1.",
    escape = FALSE
  )
mu1 <- number(as.numeric(tab_1[2, 2:4, TRUE]))
sd1 <- number(as.numeric(tab_1[4, 2:4, TRUE]))
mu1n <- tab_1[2, 2, TRUE]
sd1n <- tab_1[4, 2, TRUE]
```

@@

```{r descriptors, include = FALSE}
# Heterogeneity interval
hi90_num <- mu1n + qnorm(c(0.05, .95)) * sd1n
hi90 <- paste(
  number(hi90_num, .01),
  collapse = ", "
)
hi90_low <- paste(
  number(
    tab_1[2, 3, TRUE] + qnorm(c(0.05, .95)) * 
      tab_1[4, 3, TRUE], .01
  ),
  collapse = ", "
)

@@

```{r}
#| label: fig-1
#| fig.height: 2
#| fig.width: 8
#| fig.env: "figure*"
#| fig.align: "center"
#| fig.cap: Heterogeneity distribution of valence effects and various descriptions of their expected heterogeneity as estimated with Model 1. A. The normal density curve defined by the point estimates of the valence effect distribution's mean ($\beta_1$) and standard deviation ($\tau_1$). Shaded areas represent areas under the normal curve within 1 (dark) and 2 (light) standard deviations of the mean. B. The 90\% Heterogeneity Interval as represented by a line segment with arrows, and the dark shaded area. C. Proportion of negative valence effects in the population (dark). D. Proportion of valence effects in the population that are within the region of practical equivalence to zero (ROPE; dark).

fig_0 <- ggplot() +
  aes(xdist = dist_normal(mu1n, sd1n)) +
  scale_x_continuous(breaks = extended_breaks()) +
  coord_cartesian(xlim = c(-0.55, 0.25)) +
  theme(
    axis.title.y = element_blank(),
    axis.title.x = element_text(lineheight = rel(0.5))
  ) +
  scale_fill_manual(values = c("grey90", "grey80"), na.value = "white")
d1 <- dist_normal(mu1n, sd1n) |>
  mean_qi(.width = c(.66, .9))

fig_1 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(level)),
    .width = c(.68, .95),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$\\tau_1$"),
    x = mu1n + .05, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n + sd1n, y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  geom_text(
    label = TeX("$\\beta_1$"),
    x = mu1n - .05, y = .075,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = mu1n, xend = mu1n, y = .1, yend = 0,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed")
  ) +
  theme(axis.title.y = element_text(angle = 90))

fig_2 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(level)),
    .width = c(.90, 1),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$HI_{90}$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  ) +
  annotate(
    "segment",
    x = d1$.lower[2], xend = d1$.upper[2], y = .1, yend = .1,
    linewidth = .35,
    arrow = arrow(length = unit(4, "pt"), type = "closed", ends = "both")
  )

fig_3 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(x < 0)),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$p^d$"),
    x = mu1n, y = .175,
    size = 3,
    parse = TRUE
  )

fig_4 <- fig_0 +
  stat_slab(
    aes(fill = after_stat(between(x, -.1, .1))),
    p_limits = c(.000001, .999999),
    slab_color = "grey10",
    slab_linewidth = .4
  ) +
  geom_text(
    label = TeX("$p^{ROPE}$"),
    x = 0, y = .07,
    size = 3,
    parse = TRUE
  )

(fig_1 | fig_2 | fig_3 | fig_4) &
  scale_y_continuous(
    expand = expansion(c(0.0, 0.05))
  ) &
  labs(
    y = "Density",
    x = "Valence effect"
  ) &
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    legend.position = "none"
  ) &
  plot_annotation(tag_levels = "A")
```

The point estimates in Table \@ref(tab:tbl-fit-1) define the expected normal distribution of valence effects in the population, visualized in Figure \@ref(fig:fig-1)A. As such, Normal(`r mu1`, `r sd1`^2^) is a valid description of heterogeneity in valence effects. However, it is an incomplete description of heterogeneity for two reasons. First, it does not necessarily communicate the degree of heterogeneity in clear and actionable terms. Second, it ignores the uncertainty with which the heterogeneity parameters are estimated.

@@

To calculate a heterogeneity interval, we first specify the desired probability limits. For a 90% interval, we use the .05 and .95 percentiles (which define the central 90% of the distribution). Then, we pass those percentiles and the estimated mean and standard deviation (see Table \@ref(tab:tbl-fit-1)) to the normal quantile function ($q$; `qnorm()` in R) to obtain the interval: $HI_{90} = q([.05, .95], \beta_1, \tau_1) = q([.05, .95],$ `r mu1[1]`, `r sd1[1]`) = [`r hi90`]. In words, this equation calculates the 0.05 and 0.95 quantiles of the normal distribution defined by the mean's ($\beta_1$) and standard deviation's ($\tau_1$) point estimates: We expect 90% of valence effects in the population to fall within this interval. We illustrate this interval in Figure \@ref(fig:fig-1)B.

@@

For many applications, quantities such as proportions of slopes above or below some critical value, or within some critical range, might serve as more informative summaries. For example, we might ask "What proportion of individuals in the population respond faster to positively valenced words?" Thus, we first define a critical value, here zero. We then pass that value and the valence effect's estimated mean and standard deviation to the cumulative normal distribution function ($p$; `pnorm()` in R) to obtain the probability density in this distribution that falls below zero: $p^d = p(0, \beta_1, \tau_1) = p(0,$ `r mu1`, `r sd1`) = `r p_less_zero`. Specifically, this metric indicates the proportion of individuals in the population with negative effects. In words, we have estimated that `r p_less_zero` of the population respond faster to positively valenced words. We show this area under the curve in Figure \@ref(fig:fig-1)C.

However, using zero as a critical value might not be sufficiently informative especially when theory allows defining a smallest effect sizes of interest, or what is known as a region of practical equivalence (to zero) [@anvariUsingAnchorbasedMethods2021; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014]. For example, we might know from theory that valence effects in the interval [-0.1, 0.1] are practically equivalent to zero. To calculate, we can again use the cumulative normal distribution function ($p$) to calculate the proportion of individuals in the population whose valence effect falls within this interval or region of practical equivalence: $p^{ROPE} = p(0.1, \beta_1, \tau_1) - p(-0.1, \beta_1, \tau_1)$ = `r p_rope`. In words, `r p_rope` of the population is expected to have valence effects that are practically equivalent to zero (within the [-0.1, 0.1] interval). We visualize this area under the curve in Figure \@ref(fig:fig-1)D.

@@

Although informative, the expected normal distribution of valence effects, or any transformation of those point estimates such as the ones illustrated in Figure \@ref(fig:fig-1), ignores the uncertainty inherent in the estimated parameters, and therefore cannot communicate the analyst's uncertainty about the degree of heterogeneity. That is, the uncertainty with which we have estimated the population parameters, indicated by e.g. the confidence intervals in Table \@ref(tab:tbl-fit-1), is not carried forward to the calculations described above. For example, we only presented the expected heterogeneity interval, but not the range of credible values that it might take. For instance, if we re-calculate $HI_{90}$ based on the lowest 95%CI bounds in Table \@ref(tab:tbl-fit-1), the result is a heterogeneity interval that indicates 90% of the population's slopes falling in the [`r hi90_low`] interval. In contrast to the first interval, this interval implies much less heterogeneity and that 90% of the population's slopes are expected to fall below zero.

@@

We think that probabilistic (i.e. "bayesian") methods are uniquely applicable to estimating and describing heterogeneity in psychological phenomena in ways that naturally incorporate information about uncertainty because they deliver distributions of plausible values that could underlie hypothetical data generating processes [@gelmanBayesianDataAnalysis2013; @kruschkeBayesianDataAnalysis2017; @kruschkeDoingBayesianData2014; @mcelreathStatisticalRethinkingBayesian2020]. Moreover, these methods are widely available in modern software packages: For typical scenarios, probabilistic models are as easily estimated as their maximum likelihood counterparts.

```{r fit1-brms}
fit_1 <- brm(
  rt ~ 1 + valence + (1 + valence | person),
  data = dat,
  file = "models/brm-fit-1",
  control = list(adapt_delta = .95)
)
```

```{r}
#| label: ppcheck-1
#| eval: false

# Quick model checking
pairs(fit_1$fit, pars = variables(fit_1)[1:6])
mcmc_plot(fit_1, type = "rhat_hist")
pp_check(fit_1)
```

tab1 <- parameters(fit_1, centrality = "mean", effects = "all") |>
  tibble() |>
  select(Parameter, Mean, starts_with("CI_"))
```

@@

post1 <- as.data.frame(fit_1, variable = c("b_", "sd_", "cor_", "sigma"), regex = TRUE) |>
  tibble() |>
  rownames_to_column("Sample")
post1_sum <- fit_1 |>
  parameters(effects = "all", centrality = "mean", dispersion = TRUE) |>
  tibble() |>
  mutate(
    Mean = number(Mean, .01),
    SD = number(SD, .001)
  ) |>
  select(Parameter, Mean, SD) |>
  pivot_longer(c(Mean, SD), names_to = "Sample") |>
  pivot_wider(names_from = Parameter, values_from = value)

set.seed(191)
post1 |>
  slice_sample(n = 6) |>
  arrange(as.integer(Sample)) |>
  mutate(
    across(-Sample, ~ number(., .01))
  ) |>
  add_row(post1_sum) |>
  setNames(c("Sample", m1_parnames)) |>
  apa_table(
    span_text_columns = FALSE,
    font_size = "small",
    midrules = 6,
    caption = "Six random samples from Model 1's parameters' posterior distributions, and their summaries.",
    escape = FALSE,
    placement = "h"
  )
```

The goal of bayesian inference is the multivariate posterior probability distribution of the model's parameters. However, because closed-form solutions are not available for posterior distributions of many important types of statistical models, in practice bayesian methods rely on drawing many random samples from the posterior distribution [@gelmanBayesianDataAnalysis2013; @ravenzwaaijSimpleIntroductionMarkov2016]. Therefore, "estimate" means a large number of random samples from the posterior distribution. These samples can then be used to calculate, summarize, and visualize the posterior distribution of any desired quantity.
